---
title: ""
subtitle: Two-sample T-test on Mango Circumference
output:
  html_document:
    df_print: paged
  html_notebook: default
---

## Executive Summary

The conventional belief is that thicker mangoes taste better. This report is based on a dataset comprising 67 observations, collected from the Fruit Collection Center, FAMA Perlis, Malaysia, which includes records on the sizes and grades of Harumanis Mangoes. The manual grading process for Harumanis mangoes is carried out by highly experienced experts known for their accuracy. We are interested in whether mangoes with higher quality (Grade A) are actually thicker than those with lower grading (Grade B).

This analytics report presents results of a 2-sample t-test on circumferences between Grade A and Grade B Harumanis mangoes from the dataset. So far, the results indicate strong statistical significance, which supports the conventional belief.


Steps are summarised as follows:

1) Setup

2) Data Wrangling: 

- Original dataset description 

- Research interest and variables of interest introduction

- Data Assessment

- Data preparation/cleaning (based on variables of interest and assessment)

3) Summary Statistics on cleaned data

4) Hypothesis Testing:

- Research question

- Null Hypothesis and Alternative Hypothesis

- Assumptions and Check on Assumption 

- Testing and Test statistics

6) Conclusion



## Setup
#### Packages

```{r message = FALSE}
library(dplyr)
library(readr)
library(magrittr)
library(car)
```


#### Function

```{r}
# Invalid data: NAs and negative and Zero values
# Below function returns index of invalid data
detect_invalid_values <- function(data_vector) {
  invalid_indexes <- which(
    is.na(data_vector) | (data_vector <= 0) 
  )
  return(invalid_indexes)
}
```

#### Reading Data

```{r message = FALSE}
mango <- read_csv("Mango.csv")
```

## Data Wrangling


#### Data Description

Malaysia is renowned for its Harumanis Mango, a prized export. The grading of Harumanis Mangoes is performed by skilled workers who assess those mangoes based on their physical appearance: size, shape, and skin color. This dataset is obtained from [Kaggle](https://www.kaggle.com/datasets/mohdnazuan/harumanis-mango-physical-measurement), it records grading (A and B) and size-dependent characteristics (weight, length, and circumference) of 67 Harumanis Mangoes, picked within a 24-hour timeframe.

67*5 is the dimension of this dataset, with detailed variable description and first 10 observations presented below:

-  **No** `numeric`: Index of observations 

-  **Weight** `numeric`: Weight of mangoes in grams

-  **Length** `numeric`: Length of mangoes in centimeters

-  **Circumference** `numeric`: Circumference of mangoes in centimeters

-  **Grade** `factor`: Grade of mangoes, with 2 values: A and B


```{r prompt=TRUE, echo=FALSE}
mango %>% as_tibble()
```


#### Research Interest

The conventional belief is that thicker mangoes taste better. However, in this study, where Grade A is considered of superior quality and retailed for higher price than Grade B, our interest lies in determining if Grade A mangoes are indeed thicker (with greater circumference) than Grade B mangoes.


#### Data Assessment

Circumference and Grade, our two focal variables, are thoroughly examined below.

-  **Levels** for `Grade`: Using `baseR::unique()` function, two values, A and B, are identified, they correspond to Grade A and Grade B mangoes.  

```{r promtp=TRUE}
mango$Grade %>% unique()
```

-  **Invalid Data** for `Circumference` (NAs/Negatives/Zeros): It seems that we have no invalid data. 

```{r}
detect_invalid_values(mango$Circumference) %>% length()
```

-  **Outliers**: The Interquartile Range criterion, utilizing `baseR::boxplot()`, is employed to detect outliers. The analysis reveals one outlier among Grade A mangoes.

```{r}
boxplot <- mango %>% boxplot(Circumference ~ Grade, data = ., ylab = "Circumference")
```


#### Data Preparation

There is no invalid data to remove. We will proceed by selecting the variables of interest, converting Grade variable into a factor, and eliminating the outlier detected in `boxplot` from our dataset. First 10 observations are presented below:


```{r}
mango <- mango %>%
  # Retain 2 variables: Grade and Circumference
  select(Grade, Circumference) %>%
  # Factorize Grade variable
  mutate(Grade = factor(Grade, levels = c("A", "B"), ordered = TRUE)) %>%
  # Filter out outliers
  filter(!(Circumference %in% boxplot$out))
```

```{r echo=FALSE}
mango %>% tibble()
```


## Summary Statistics

From the descriptive statistics, it's clear that the sample mean for Grade A mangoes is significantly greater than that of Grade B mangoes, with equal sample sizes for both groups. Additionally, Grade A does not show a significantly smaller interquartile range than Grade B.

```{r}
mango %>% 
  group_by(Grade) %>%  # Group by Grade
  summarise(
    # Minimum value
    min = min(Circumference, na.rm = TRUE), 
    # Maximum value
    max = max(Circumference, na.rm = TRUE), 
    # Mean
    mean = mean(Circumference, na.rm = TRUE),
    # Standard deviation
    stand_dev = sd(Circumference, na.rm = TRUE), 
    # First quantile
    first_quantile = quantile(Circumference, 0.25, na.rm = TRUE),
    # Third quantile
    third_quantile = quantile(Circumference, 0.75, na.rm = TRUE),
    # Interquartile range
    iqr = IQR(Circumference, na.rm = TRUE),
    # Number of observations
    n = n(),
    # Probability by each Grade
    prob = n()/length(mango$Grade)
  )

```

## Hypothesis Testing

#### Research Question

> Whether Grade A mangoes possess greater circumference than Grade B mangoes?

#### Null Hypothesis 

Population mean of circumference for Grade A mangoes will be equal to that of Grade B mangoes

$$
H_0: \mu_A - \mu_B =0
$$

#### Alternative Hypothesis 

Population mean of circumference for Grade A mangoes will be greater than that of Grade B mangoes

$$
H_0: \mu_A - \mu_B > 0
$$

#### Assumptions and Significant level

- Data from our samples follows a normal distribution.

- Sampling distributions of the mean for Grade A and Grade B mangoes have equal variances.

- Chosen significance level is 0.05.

#### Assumptions' Tests

- **Normality Check**

QQ plots are used to assess the normality of our sample data's distribution. When examining the QQ plots for Grade A and Grade B mangoes, the data points closely follow a straight line, and fall within the shaded region of 95% confidence level for normal distribution. Hence, it is safe to proceed with the assumption of normal distribution

```{r results='hide'}
# Subsetting into Grade A mangoes dataset
gradeA <- mango %>%
  filter(Grade == "A")

# Plotting
gradeA$Circumference %>% qqPlot(dist="norm",
                                envelope=list(level=0.95))
```


```{r results='hide'}
# Subsetting into Grade B mangoes dataset
gradeB <- mango %>%
  filter(Grade == "B")

# Plotting
gradeB$Circumference %>% qqPlot(dist="norm",
                                envelope=list(level=0.95))
```


- **Homogeneity of Variance Test**

*Null Hypothesis*

$$
H_0: \mu_A - \mu_B =0
$$
Homogeneity of Variance assumption is tested with Levene Test, by using `leveneTest()` function in R. Since `Pr(>F)` p-value is 0.15, greater than 0.05 - our chosen significant level, this test is not statistically significant. Therefore, it is safe to assume the two populations having the same variance.

```{r}
mango %>% leveneTest(Circumference ~ Grade, data =.)
```

#### Testing

We use `t.test()` function from base R to perform a Welch two-sample right-tailed t-test with a significance level of 0.05 or confidence level of 0.95

```{r}
test <- mango %>% 
  t.test(
    Circumference ~ Grade,
    data = .,
    # Assuming equal variance between 2 populations
    var.equal = TRUE,
    # Right-tailed t-test
    alternative = "greater",
    # Confidence level = 1 - Rejection Region/Significant Level
    conf.level = 0.95)
```
```{r echo=FALSE}
test
```

#### Statistics

- **Test Statistics**

```{r}
data.frame(
  "test statistics" = test$statistic,
  "sample mean difference" = mean(gradeA$Circumference) - mean(gradeB$Circumference)
)
```


- **Critical Mean Difference**

Under the null hypothesis and assumption of equal variance, sampling distribution of the difference in circumference between Grade A and Grade B mangoes follows a t-distribution with mean = 0 and 64 degrees of freedom. Since this is a right-tailed t-test, critical value to mean difference will be located in the upper quantile, and be calculated as follow

```{r}
# df = df1 + df2 - 2 
# Where df1 is number of observations for Grade A mangoes, df2 is number of observations for Grade B mangoes
qt(p = 0.05, df = 33 + 33 - 2, lower.tail = FALSE)
```

- **P-value**

Assuming there is no difference in circumference between Grade A and Grade B mangoes, the p-value represents the probability of obtaining a sample where the average circumference of Grade A is 1.872 centimeters greater than that of Grade B under this assumption 

$$
p-value = Pr(\bar{x}_A - \bar{x}_B > 1.872 | \mu_A - \mu_B = 0)
$$
Or: 
$$
p-value = Pr(t > 6.139 | t = 0)
$$

```{r}
test$p.value
```

- **Confidence Intervals**

```{r}
data.frame(
  "lower bound" = test$conf.int[1],
  "upper bound" = test$conf.int[2]
)
```


#### Drawings on Statistics

Our test is statistically significant, leading us to reject the null hypothesis based on the following observations:

- The test statistic (6.139) exceeds the critical mean difference (1.669).

- The p-value is well below the 0.05 significance level, approaching zero.

We favor the alternative hypothesis, as the confidence interval does not include 0; indeed, it is entirely above zero. Specifically, we are 95% confident that the difference between grade A and grade B population means will be greater than 1.363.

## Conclusion

Following the 2-sample t-test on Mango Circumference dataset, we can conclude that Grade A mangoes indeed have a greater circumference than Grade B mangoes. This finding aligns with the conventional belief and provides valuable guidance for informed purchasing decisions, given that the grading process is performed by experienced farmers.


## References

OpenAI (2023) ChatGPT (May 24 version) [Large language model], accessed October 12th 2023, https://chat.openai.com/c/e40b0a42-55a2-4c1a-b121-a553cd0707d3


Mohammad HI, Tajul RR, Ray AJG and NW (2022). Harumanis Mango Physical Measurements [Data set], Kaggle, accessed 10th October 2023, https://www.kaggle.com/datasets/mohdnazuan/harumanis-mango-physical-measurement


Mohd FI, Fathinul SA, Ammar Z and Ali YM (2016). In-Line Sorting of Harumanis Mango Based on External Quality Using Visible Imaging, Sensors2016, 16, 1753,1-17,doi:10.3390/s16111753, accessed 10th October 2023,
https://mdpi-res.com/sensors/sensors-16-01753/article_deploy/sensors-16-01753.pdf?version=1477567023


Mohd NW, Mohammad HI, Nurul AZ,and SF (7-8 Sept. 2022)'Grading of Mango Fruits Based on Physical Measurements'[Published Paper], 2022 3rd INTERNATIONAL CONFERENCE ON ARTIFICIAL INTELLIGENCE AND DATA SCIENCES (AiDAS),IPOH, Malaysia, accessed October 14,2023,
https://ieeexplore-ieee-org.ezproxy.lib.rmit.edu.au/stamp/stamp.jsp?tp=&arnumber=9918732
